Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-16 paper/paper.tex
Added notes to Section 9

@@

\citet{barras2010sets, Wang2013SemanticsOI} assign
set-theoretic semantics to dependent type theory in Coq. Unlike the
previous efforts, which primarily focus on predicative
type theory and more direct reducibility models,
\citet{barras2010sets, Wang2013SemanticsOI} tackle extensions of
$CC^\omega$, a system that incorporates a predicative universe on top
of the impredicative sort in the Calculus of Constructions. We choose to
focus on a syntactic term model so we do not have to take the extra step
of mechanizing mathematical objects such as sets and domains.

There are other mechanized developments for dependently typed systems that
only involve properties that are derivable through syntactic means. For
example, \citet{coqcoqcorrect2019} prove the correctness of a type checker
for the Predicative, Cumulative Calculus of Inductive Constructions (PCUIC),
Coq's core calculus, assuming the strong normalization property of the object
language. \citet{weirich:systemd} define System D, a core calculus of
dependent Haskell, and prove the syntactic type safety of the type
system. Because System D includes nontermination, they proved
the consistency of definitional equality from the confluence of 
parallel reduction.

Compared to the systems described here, the most notable features we are
missing are cumulativity and impredicativity. Our semantic model already
satisfies the cumulativity property (Lemma~\ref{lemma:logrelcumulativity}),
but we need to extend our convertibility relation into a subtyping relation in
our syntactic typing rules. Impredicativity, on the other hand, is known to be
difficult to model when the impredicative sort is at the bottom of a
predicative universe hierarchy; in this scenario, the erasure technique from
\citet{geuvers1994short} is not
applicable~\citep{abel2013normalization}. Whether there is a similarly short
and simple treatment for impredicativity remains an open question.

>>>>

\jc{Missing from the table:
\begin{itemize}
  \item Altenkirch and Kaposi (2016) using QIITs, mentioned above (Agda);
    mechanization might be part of Kaposi's thesis (2016)
  \item prior work Chapman (2009), Type Theory should Eat Itself (Agda);
    possibly also his thesis (2008)
  \item Kipling by McBride (2010), mixing deep and shallow embedding
    using host conversion (Agda) to avoid the whole QIIT business
  \item Kaposi et al. (2019), Shallow Embedding of Type Theory (Agda),
    although at this point I'm not sure if this counts as having a syntax
\end{itemize}
}

@@

\Citet{barras2010sets} and \citet{Wang2013SemanticsOI} assign
set-theoretic semantics to dependent type theory in Coq. Unlike the
previous efforts, which focus on predicative type theory and direct
reducibility models, they tackle extensions of
CC$^\omega$, which extends the Calculus of Constructions with predicative
universes on top of the impredicative sort. We choose to
focus on a syntactic term model to avoid mechanizing mathematical objects
such as sets and domains.

There are other mechanized developments for dependently typed systems that
only involve properties that are derivable through syntactic means. For
example, \citet{coqcoqcorrect2019} prove the correctness of a type checker
for the Predicative Calculus of Cumulative Inductive Constructions (pCuIC),
Coq's core calculus, assuming strong normalization.
\citet{weirich:systemd} define System D, a core calculus of
dependent Haskell, and prove the syntactic type safety of the type
system. Because System D includes nontermination, they prove
consistency of definitional equality from the confluence of 
parallel reduction.
\jc{Why specifically mention System D? Surely there are other more notable
(in reviewers' eyes) dependent languages for which type safety have been proven?}

Compared to the systems described here, the most notable feature we are
missing is impredicativity, which is known to be
difficult to model when the impredicative sort is at the bottom of a
predicative universe hierarchy.
\jc{This makes it sound like it's easy when it's at the top of a predicative
universe hierarchy, which I don't think is what you meant; did you mean when
impredicativity is combined with some sort of large elimination of base types/
type-level computation?}
In this scenario, the erasure technique from \citet{geuvers1994short} is not
applicable~\citep{abel2013normalization}. Whether there is a similarly short
and simple treatment for impredicativity remains an open question.
